Generalized spaces for constructive algebra

### 2020-04-07 paper-herrsching2019/main.tex
Lots of tiny improvements

\newcommand{\hcancel}[5]{%
  \tikz[baseline=(tocancel.base)]{
    \node[inner sep=0pt,outer sep=0pt] (tocancel) {#1};
    \draw[red!80, line width=0.4mm] ($(tocancel.south west)+(#2,#3)$) -- ($(tocancel.north east)+(#4,#5)$);
  }%
}

@@

\address{Universität Augsburg \\
Institut für Mathematik \\
Universitätsstr. 14 \\
86159 Augsburg, Germany}

@@

The notion of \emph{space} is fundamental to large parts of mathematics. This notion
exists in various flavors, ranging from the basic cartesian spaces~$\RR^n$ to
the more general metric and topological spaces and also including several more
and slightly exotic flavors such as the diffeological spaces of differential topology.

@@

% NOTE TO EDITORS AND LAYOUTERS: It is on purpose that these two definitions
% are not set using the usual theorem environment. This is because we are only
% citing these definitions in order to reflect on them. We are not actually
% introducing these two notions here for further use.
\begin{quote}
\textbf{Definition.} A \emph{metric space} consists of a set~$X$ of points
together with a metric~$d : X \times X \to \RR_{\geq 0}$ satisfying the metric
axioms.

@@

Locales are a further and particularly unique flavor of the notion of space
which turn this classical picture upside down: Locales embrace \emph{opens
instead of points as primitive building blocks}. With locales, points are a
derived concept. In particular, the opens of a locale are not sets of points;
in fact, they needn't be sets of anything in peculiar.

@@

The points of~$X$ are in canonical one-to-one correspondence with the
surjections~$\NN \twoheadrightarrow \RR$, hence~$X$ does not have any points.
But this locale does have uncountably many basic opens~$U_{nx}$, where~$n$
ranges over the naturals and~$x$ ranges over the reals. We picture~$U_{nx}$ as
the ``open of those surjections~$f$ for which~$f(n) = x$'', we can compute with
these opens, consider functions on~$X$ and ponder sublocales of~$X$. For instance:
\begin{enumerate}
\item If~$x \neq y$, then the intersection of~$U_{nx}$ with~$U_{ny}$
is truly empty.
\item The union of the~$U_{nx}$, where~$x$ is a fixed real
number and~$n$ ranges over the naturals, is all of~$X$. No finite number
of these opens covers~$X$, hence~$X$ is not compact.
\item For any real~$x$, there is a well-defined continuous function from~$X$ to
(the localic version of) the naturals which, on the level of points, would map
a surjection~$f : \NN \to \RR$ to the smallest preimage of~$x$ under~$f$.
\item There is a well-defined continuous function from~$X$ to (the localic
version of) the reals which, on the level of points, would map a surjection~$f$
to the number~$\sum_{n=0}^\infty 2^{-n} \arctan(f(n))$. Unlike the previous
example, this function is not locally constant.
\item\label{item:intersection-sublocales} The locale~$X$ is a sublocale of the
locale~$Y$ of arbitrary functions~$\NN \to \RR$ (which can also be realized as
a topological space). It can be obtained as the intersection of the uncountably
many sublocales~$Y_x$, where~$Y_x$ is the sublocale of~$Y$ consisting of those
functions which hit the real~$x$.
\item A certain sublocale of~$X$, the locale~$X'$ of those functions~$\NN \to
\RR$ for which any real has infinitely many preimages, has a fractal nature: It
is covered by the (intersections with~$X'$ of the)~$U_{0x}$, where~$x$ ranges over the reals; the pairwise
intersection of these opens is (truly) empty; and they are each isomorphic
to~$X'$ -- on the level of points, by mapping a surjection~$f$ in~$U_{0x}$ to
the surjection~$(n \mapsto f(n+1))$.
\end{enumerate}

@@

\paragraph{The Banach--Tarski paradox.} The Banach--Tarski paradox is the
unintuitive statement that a three-dimensional solid ball in~$\RR^3$ of
radius~$r$ can be partitioned into six disjoint subsets in such a way that
rearranging those subsets using only Euclidean motions yields two disjoint
solid balls of radius~$r$ each. The axiom of choice is required to construct
these subsets, and the Banach--Tarski paradox is not in
contradiction with the basic properties of the Lebesgue measure in~$\RR^3$
because these intermediate subsets are not measurable.

@@

The Banach--Tarski paradox can also be avoided by adopting a localic point of
view: While the localic counterparts of the six pieces do not have any points
in common, the locale-theoretic pairwise intersections are still nontrivial~\cite{simpson:measure}.
Hence there is no paradox, as one would not expect an rearrangement of
overlapping sublocales to preserve volume.

@@

Unlike a regular sequence such as~$(0,0,0,\ldots)$, we expect that, if the
digits of a \emph{random sequence} are revealed step by step, we will
eventually verify that it is an element of~$W$; and moreover, that it has any
observable property~$U$ with Lebesgue measure~$1$.

@@

By relinquishing points, locales provide a more flexible notion of space.
Section~\ref{sect:examples-no-points} substantiates this observation
with several examples of nontrivial locales without any points. Constructive
mathematics gives a further, orthogonal motivation to study locales: There are
situations in which the relevant spaces do have enough points, but only if one
subscribes to the axiom of choice or similar non-constructive principles. In
these situations, the pointfree approach allowed by locales helps to give
constructive versions of classical results. We will further discuss this theme
in Section~\ref{sect:presenting-frames}.

XXX ideal objects

@@

\paragraph{Galois theory.} Let~$L|k$ be a Galois extension. The
fundamental theorem of Galois theory states that there is a bijection between
the intermediate extensions~$L|E|k$ and the closed subgroups of the topological
Galois group~$\Gal(L|k)$. The bijection maps an intermediate extension~$L|E|k$
to the subgroup~$\Gal(L|E)$ and its inverse maps a closed subgroup~$H \subseteq
\Gal(L|k)$ to the intermediate extension~$L^H$.

@@

The point~$x$ is completely determined by the
set~$\{ U \in \O(X) \,|\, \alpha(U) = \top \}$. This set is a \emph{completely
prime filter}, that is a subset~$\fff \subseteq \O(X)$ which is upward-closed,
closed under finite meets and for which~$\bigvee_i U_i \in \fff$ implies~$U_i
\in \fff$ for some index~$i$; and conversely, any such completely prime filter
gives rise to a point of~$X$ (Exercise~\ref{ex:points-as-filters}).

\begin{definition}A locale~$X$ is \emph{spatial} if and only if its points
suffice to distinguish its opens, that is if for any opens~$U, V \in \O(X)$,
if~$x \inplus U \Rightarrow x \inplus V$ for all points~$x$ of~$X$, then~$U
\preceq V$.\end{definition}

An equivalent conditon is: if for any opens~$U, V \in \O(X)$, if~$x \inplus U
\Leftrightarrow x \inplus V$ for all points~$x$ of~$X$, then~$U = V$.

@@

\begin{remark}\label{rem:sober-embed}
Since continuous maps between spaces induce locale morphisms, we
have a canonical map~$\Hom_\Top(Y,Y') \to \Hom_\Loc(L(Y),L(Y'))$. If~$Y'$ is
sober, then this map is a bijection. In particular, in this case the set-theoretic points
of~$Y'$ and the locale-theoretic points of~$L(Y')$ are in canonical
one-to-one correspondence.
\end{remark}

\subsection{Exercises}

@@

\begin{exercise}{Embedding sober topological spaces into locales}%
Verify the statements made in Remark~\ref{rem:sober-embed}.
\end{exercise}

\begin{exercise}{Hausdorff spaces are sober}%
\label{ex:hausdorff-sober}%
A \emph{meet-irreducible open subset} of a topological space~$Y$ is an
open subset~$W \subseteq Y$ such that~$(U \cap V \subseteq W) \Rightarrow (U
\subseteq W) \vee (V \subseteq W)$ for all open subsets~$U,V \subseteq Y$ and
such that~$Y \not\subseteq W$.
\begin{alphlist}[(c)]
\item Assuming the law of excluded middle, show that the completely prime
filters of a topological space are in canonical one-to-one correspondence with
the meet-irreducible open subsets.\smallskip

@@

\item Assuming the law of excluded middle, show that Hausdorff topological
spaces validate the condition stated in part~(b) and hence are sober.
\end{alphlist}
{\scriptsize\emph{Note.} Without the law of excluded middle, Hausdorff spaces
and even metric spaces may fail to be sober. In fact, already the topological
space of the rational numbers can fail to be
sober~\cite[Example~8.14(iv)]{fourman-scott:sheaves-and-logic}.\par}
\end{exercise}

\begin{exercise}{Complete metric spaces are sober}
A \emph{metric space} is a set~$Y$ together with a relation~$d \subseteq Y
\times Y \times \QQ_{\geq0}$, written~``$d(x,y) \leq q$'', such that the
following axioms are satisfied.
\begin{align*}
  d(x,y) \leq q \wedge q \leq r &\Longrightarrow d(x,y) \leq r \\
  (\forall r \in \QQ_{> q}\_ d(x,y) \leq r) &\Longrightarrow d(x,y) \leq q \\
  &\mathrel{\phantom{\Longrightarrow}} \exists q \in \QQ_{\geq0}\_ d(x,y) \leq q \\
  &\mathrel{\phantom{\Longrightarrow}} d(x,x) \leq 0 \\
  d(x,y) \leq 0 &\Longrightarrow x = y \\
  d(x,y) \leq q &\Longrightarrow d(y,x) \leq q \\
  d(x,y) \leq q \wedge d(y,z) \leq r &\Longrightarrow d(x,z) \leq q+r
\end{align*}
A \emph{Cauchy process} (or \emph{multi-valued modulated Cauchy sequence}) in a metric
space~$Y$ is a map~$\alpha : \QQ_{>0} \to P(Y)$ such that all sets~$\alpha(q)$
are inhabited and such that~$d(x,x') \leq q + q'$ for all~$q,q' \in \QQ_{>0}$
and~$x \in \alpha(q), x' \in \alpha(q')$. Such a process \emph{converges
to}~$x_0 \in Y$ iff~$d(x,x_0) \leq q$ for all~$q \in \QQ_{>0}$
and~$x \in \alpha(q)$. A metric space~$Y$ is \emph{complete} iff every Cauchy
process in~$Y$ converges.

Verify, without using the law of excluded middle, that any complete metric
space is sober.
\end{exercise}

\begin{exercise}{Isomorphisms of frames}%
Show that a frame homomorphism~$\alpha : A \to A'$ is an isomorphism (that is,
admits an inverse frame homomorphism) if and only if~$\alpha$ is surjective and
reflects the ordering (that is,~$\alpha(U) \preceq \alpha(V) \Rightarrow U
\preceq V$).
\end{exercise}

\begin{exercise}{The covariant approach to locales}%
Let~$A$ be a frame.
\begin{alphlist}[(c)]
\item Show that~$A$ contains arbitrary (set-indexed) meets (though they are not
required to be preserved by frame homomorphisms), by the impredicative
construction
\[ \bigwedge_{i \in I} U_i = \bigvee\{ \text{$V \preceq U_i$ for all~$i \in I$}
\}. \]
\item What are the arbitrary meets in the frame of open subsets of a
topological space?
\item Show that for any frame homomorphism~$\alpha : A \to A'$ there exists a
monotone map~$\beta : A' \to A$ such that~\emph{$\beta$ is right adjoint
to~$\alpha$}, that is such that
\[ \alpha(U) \preceq V \quad\text{iff}\quad U \preceq \beta(V) \]
for all~$U \in A$ and~$V \in A'$.
\end{alphlist}
{\scriptsize\emph{Note.} Based on the observation in part~(c), a
\emph{covariant approach} to locales can be developed, in which locale maps~$X
\to X'$ are defined to be certain kinds of maps~$\O(X) \to
\O(X')$~\cite{picado-pultr:covariant}.\par}
\end{exercise}

@@

@@

\begin{exercise}{Coproducts of locales}%
\begin{alphlist}[(c)]
\item Show that the category of frames has small products.
\item Deduce that the category of locales has small coproducts.
\item Let~$M$ be a set and endow it with the discrete topology. Show
that~$L(M)$ is the coproduct~$\coprod_{x \in M} \pt$.
\end{alphlist}
{\scriptsize\emph{Note.} The category of locales also has small products;
Exercise~\ref{ex:products-of-locales} is devoted to this fact.\par}
\end{exercise}

@@

\section{Sheaf models}

@@

\begin{definition}A \emph{geometric theory} consists of
\begin{enumerate}
  \item a set of sorts: $X$, $Y$, $Z$, \ldots
  \item a set of function symbols: $f : X \times Y \to Z$, \ldots
  \item a set of relation symbols: $R \hookrightarrow X \times Y \times Z$, \ldots
  \item a set of geometric sequents as axioms: $\varphi \vdash_{x:X, y:Y} \psi$, \ldots
\end{enumerate}
A \emph{geometric sequent} (in some context~$x_1\?X_1,\ldots,x_n\?X_n$) is a
formula built using only the ingredients ${=}\ {\top}\ {\wedge}\ {\bot}\
{\vee}\ {\bigvee}\ {\exists}$ and the relation symbols (but no ${\Rightarrow}\
{\forall}$). The symbol~``$\bigvee$'' refers to disjunction of arbitrary
set-indexed families of formulas. The theory \emph{proves} a sequent~$\sigma$ if and
only if there exists a \emph{derivation} of~$\sigma$, where the set of
derivations is inductively generated by the axioms and the rules displayed in
Table~\ref{table:geometric-logic}.\footnote{Because of the infinitary
disjunctions, some care is needed to not ``built in'' the axiom of choice in
the definition of derivations. More specifically, we allow inhabited sets of
subdeductions instead of requiring that subdeductions are given by a function.
Details on the implementation in~\textsc{czf} can be found in
Ref.~\cite[Definition~5.2]{rathjen:barr}.}
\end{definition}

@@

    (structural rules and rules for~${=}\ {\top}\ {\wedge}\ {\bot}\ {\vee}\
    {\exists}$ as in Table~\ref{table:geometric-logic})
    \medskip

@@

Even though superficially similar, geometric theories play a
substantially different role than formal systems such as Peano arithmetic or
Zermelo--Fraenkel set theory. There are a number of notable differences:
\begin{enumerate}
\item Formal systems are typically of foundational interest and can be
fruitfully employed as metatheories. Geometric theories are more interesting
for applications, for instance for presenting spaces. This aspect is discussed
in Section~\ref{sect:presenting-frames}.
\item Formal systems typically come with an intended model; geometric theories
do not. For instance, the geometric theory of groups has all groups as models;
when we are writing down the axioms of a group, we are not setting out to
capture the properties of any one specific group.
\item Formal systems are typically required to be recursively
axiomatizable. Many interesting geometric theories are not, and indeed contain
an uncountable number of sorts, function symbols, relation symbols or axioms. Geometric theories also allow for arbitrary
set-indexed disjunctions; for this to make sense, the study of geometric
theories can only be carried out in a sufficiently rich background theory.
\item For formal systems, typically a version of Gödel's completeness theorem
holds: A formula is derivable iff it holds in all models. In contrast,
geometric theories can be consistent yet do not admit set-based models. An
example is the geometric theory of surjections~$\NN \twoheadrightarrow \RR$
reviewed below.
\item Perhaps the most important difference to formal systems is that geometric
theories often depend on given mathematical objects. For instance, for each
ring~$A$, there is the geometric theory of prime ideals of~$A$. This theory
will play a crucial role in Section~\ref{sect:appl}.
\end{enumerate}

@@

In a similar vein, there is the geometric theory of monoids, of groups, and so
on. Some of these theories use the infinite disjunctions supported by geometric
logic: For instance, the theory of fields of positive characteristic or the
theory of torsion groups have among their axioms
\[ \top \vdash \bigvee_{n \in \NN_{\geq 1}} (\underbrace{1 + \cdots + 1}_{\text{$n$ summands}} = 0)
  \qquad
  \top \seq{g:G} \bigvee_{n \in \NN} (g^n = e). \]

@@

\item \emph{The geometric theory of automorphisms of a field.}\label{par:theory-automorphisms}
Let~$L|k$ be a field
extension. The classifying locale of automorphisms~$L \to L$ which fix~$k$ has
one nullary relation symbol~``$\sigma_{xy}$'' for each pair~$\langle x,y
\rangle \in L$ and the following axioms:
\begin{align*}
  \top &\vdash \bigvee_y \sigma_{xy} & \text{(for each~$x \in L$)} \\
  \top &\vdash \bigvee_x \sigma_{xy} & \text{(for each~$y \in L$)} \\
  \sigma_{xy} \wedge \sigma_{xy'} &\vdash \bigvee\{ \top \,|\, y = y' \} & \text{(for each~$x,y,y' \in L$)} \\
  \sigma_{xy} \wedge \sigma_{x'y} &\vdash \bigvee\{ \top \,|\, x = x' \} & \text{(for each~$x,x',y \in L$)} \\
  \top &\vdash \sigma_{00} \\
  \top &\vdash \sigma_{11} \\
  \sigma_{xy} \wedge \sigma_{x'y'} &\vdash \sigma_{x+x',y+y'} & \text{(for each~$x,x',y,y' \in L$)} \\
  \sigma_{xy} \wedge \sigma_{x'y'} &\vdash \sigma_{xx',yy'} & \text{(for each~$x,x',y,y' \in L$)} \\
  \top &\vdash \sigma_{xx} & \text{(for each~$x \in k$)}
\end{align*}
A model of this theory (can be reorganized to be) an element of the classical
Galois group of~$L|k$.

@@

\subsection{Sheaves on locales}

@@

>>>>

\newcommand{\hcancel}[5]{%
  \tikz[baseline=(tocancel.base)]{
    \node[inner sep=0pt,outer sep=0pt] (tocancel) {#1};
    \draw[black, line width=0.2mm] ($(tocancel.south west)+(#2,#3)$) -- ($(tocancel.north east)+(#4,#5)$);
  }%
}

@@

\address{Università di Verona \\
Department of Computer Science \\
Strada le Grazie 15 \\
37134 Verona, Italy}

@@

The notion of \emph{space} is fundamental to large parts of mathematics. This notion
exists in various flavors, ranging from the basic cartesian spaces~$\RR^n$ to
the more general metric and topological spaces and also including several more
and slightly exotic flavors such as the diffeological spaces from differential topology.

@@

% NOTE TO EDITORS AND LAYOUTERS: It is on purpose that these two definitions
% are not set using the usual theorem environment. This is because we are only
% citing these definitions in order to reflect on them. We are not actually
% introducing these two notions here for further use.
\begin{quote}
\textbf{Definition.} A \emph{metric space} consists of a set~$X$ of points
together with a map~$d : X \times X \to \RR_{\geq 0}$ satisfying the metric
axioms.

@@

Locales are a further and particularly unique flavor of the notion of space
which turn this classical picture upside down: Locales embrace \emph{opens
instead of points as primitive building blocks}. With locales, points are a
derived concept. In particular, the opens of a locale are not sets of points;
in fact, they need not be sets of anything in peculiar.

@@

The points of~$X$ are in canonical one-to-one correspondence with the
surjections~$\NN \twoheadrightarrow \RR$, hence~$X$ does not have any points.
But this locale does have uncountably many basic opens~$U_{nx}$, where~$n$
ranges over the naturals and~$x$ ranges over the reals. We picture~$U_{nx}$ as
the ``open of those surjections~$f$ for which~$f(n) = x$'', we can compute with
these opens, consider functions on~$X$ and ponder sublocales of~$X$. For instance:
\begin{enumerate}
\item If~$x \neq y$, then the intersection of~$U_{nx}$ with~$U_{ny}$
is truly empty.
\item The union of the~$U_{nx}$, where~$x$ is a fixed real
number and~$n$ ranges over the naturals, is all of~$X$. No finite number
of these opens covers~$X$, hence~$X$ is not compact.
\item Assuming~$x = y \vee \neg(x = y)$ for any pair of reals, there is, for any real~$x$, a well-defined continuous function from~$X$ to
(the localic version of) the naturals which, on the level of points, would map
a surjection~$f : \NN \to \RR$ to the smallest preimage of~$x$ under~$f$.
\item There is a well-defined continuous function from~$X$ to (the localic
version of) the reals which, on the level of points, would map a surjection~$f$
to the number~$\sum_{n=0}^\infty 2^{-n} \arctan(f(n))$. Unlike the previous
example, this function is not locally constant.
\item\label{item:intersection-sublocales} The locale~$X$ is a sublocale of the
locale~$Y$ of arbitrary functions~$\NN \to \RR$ (which can also be realized as
a topological space). It can be obtained as the intersection of the uncountably
many sublocales~$Y_x$, where~$Y_x$ is the sublocale of~$Y$ consisting of those
functions which hit the real~$x$.
\item A certain sublocale of~$X$, the locale~$X'$ of those functions~$\NN \to
\RR$ for which any real has infinitely many preimages, has a fractal nature: It
is covered by the (intersections with~$X'$ of the)~$U_{0x}$, where~$x$ ranges over the reals; the pairwise
intersection of these opens is (truly) empty; and they are each isomorphic
to~$X'$ -- on the level of points, by mapping a surjection~$f$ in~$U_{0x}$ to
the surjection~$(n \mapsto f(n+1))$.
\end{enumerate}

@@

\paragraph{The Banach--Tarski paradox.} The Banach--Tarski paradox is the
unintuitive statement that a three-dimensional solid ball in~$\RR^3$ of
radius~$r$ can be partitioned into five disjoint subsets in such a way that
rearranging those subsets using only Euclidean motions yields two disjoint
solid balls of radius~$r$ each. The axiom of choice is required to construct
these subsets, and the Banach--Tarski paradox is not in
contradiction with the basic properties of the Lebesgue measure in~$\RR^3$
because these intermediate subsets are not measurable.

@@

The Banach--Tarski paradox can also be avoided by adopting a localic point of
view: While the localic counterparts of the five pieces do not have any points
in common, the locale-theoretic pairwise intersections are still nontrivial~\cite{simpson:measure}.
Hence there is no paradox, as one would not expect a rearrangement of
overlapping sublocales to preserve volume.

@@

Unlike a regular sequence such as~$(0,0,0,\ldots)$, we expect that, if the
digits of a \emph{random sequence} are revealed step by step, we will
eventually verify that it is an element of~$W$; and moreover, that it has any
observable property~$U$ of Lebesgue measure~$1$.

@@

By relinquishing points, locales provide a more flexible notion of space.
Section~\ref{sect:examples-no-points} substantiates this observation
with several examples of nontrivial locales without any points. Constructive
mathematics gives a further, orthogonal motivation to study locales: There are
situations in which the relevant spaces do have enough points, but only if one
subscribes to the axiom of choice or similar non-constructive principles. In
these situations, the pointfree approach facilitated by locales helps to give
constructive versions of classical results. We will further discuss this theme
in Section~\ref{sect:presenting-frames}. Examples for this phenomenon include
the following.

@@

\paragraph{Galois theory.} Let~$L|k$ be a Galois extension. The
fundamental theorem of (infinite) Galois theory states that there is a bijection between
the intermediate extensions~$L|E|k$ and the closed subgroups of the topological
Galois group~$\Gal(L|k)$. The bijection maps an intermediate extension~$L|E|k$
to the subgroup~$\Gal(L|E)$ and its inverse maps a closed subgroup~$H \subseteq
\Gal(L|k)$ to the fixed field~$L^H$.

@@

The point~$x$ is completely determined by the information to which opens it
belongs, that is by the the set {$\{ U \in \O(X) \,|\, \alpha(U) = \top \}$}. This set is a \emph{completely
prime filter}, that is a subset~$\fff \subseteq \O(X)$ which is upward-closed,
closed under finite meets and for which~$\bigvee_i U_i \in \fff$ implies~$U_i
\in \fff$ for some index~$i$; and conversely, any such completely prime filter
gives rise to a point of~$X$ (Exercise~\ref{ex:points-as-filters}).

\begin{definition}A locale~$X$ is \emph{spatial} if and only if its points
suffice to detect the inclusion relation on the opens of~$X$, that is if for any opens~$U, V \in \O(X)$,
if~$x \inplus U \Rightarrow x \inplus V$ for all points~$x$ of~$X$, then~$U
\preceq V$.\end{definition}

An equivalent definition is: A locale~$X$ is \emph{spatial} if and only if its points
suffice to distinguish its opens, that is if for any opens~$U, V \in \O(X)$, if~$x \inplus U
\Leftrightarrow x \inplus V$ for all points~$x$ of~$X$, then~$U = V$.

@@

\begin{remark}\label{rem:sober-embed}
Since continuous maps between spaces induce locale morphisms, we
have a canonical map~$\Hom_\Top(Y,Y') \to \Hom_\Loc(L(Y),L(Y'))$. If~$Y'$ is
sober, this map is a bijection. In particular, in this case the set-theoretic points
of~$Y'$ and the locale-theoretic points of~$L(Y')$ are in canonical
one-to-one correspondence.
\end{remark}

\subsection*{Exercises}
\addcontentsline{toc}{subsection}{Exercises}

@@

\begin{exercise}{Sober topological spaces as locales}%
Verify the statements made in Remark~\ref{rem:sober-embed}.
\end{exercise}

% FUTURE: Fix linebreak
\begin{exercise}{Hausdorff spaces are sober}%
\label{ex:hausdorff-sober}%
A \emph{meet-irreducible open subset} of a topological space~$Y$ is an
open subset~$W \subseteq Y$ such that~$(U \cap V \subseteq W) \Rightarrow (U
\subseteq W) \vee (V \subseteq W)$ for all open subsets~$U,V \subseteq Y$ and
such that~$Y \not\subseteq W$.
\begin{alphlist}[(c)]
\item Assuming the law of excluded middle, show that the completely prime
filters of a topological space are in canonical one-to-one correspondence with
the meet-irreducible open subsets.\smallskip

@@

\item Assuming the law of excluded middle, show that Hausdorff topological
spaces validate the condition stated in part~(b) and hence are sober.
\end{alphlist}
{\scriptsize\emph{Note.} Without the law of excluded middle, Hausdorff spaces
and even metric spaces may fail to be sober. In fact, already the topological
space of rational numbers can fail to be
sober~\cite[Example~8.14(iv)]{fourman-scott:sheaves-and-logic}.\par}
\end{exercise}

\begin{exercise}{Complete metric spaces are sober}
A \emph{metric space} is a set~$Y$ together with a relation~$d \subseteq Y
\times Y \times \QQ_{\geq0}$, written~``$d(x,y) \leq q$'', such that the
following axioms are satisfied.
\begin{align*}
  d(x,y) \leq q \wedge q \leq r &\Longrightarrow d(x,y) \leq r \\
  (\forall r \in \QQ_{> q}\_ d(x,y) \leq r) &\Longrightarrow d(x,y) \leq q \\
  &\mathrel{\phantom{\Longrightarrow}} \exists q \in \QQ_{\geq0}\_ d(x,y) \leq q \\
  &\mathrel{\phantom{\Longrightarrow}} d(x,x) \leq 0 \\
  d(x,y) \leq 0 &\Longrightarrow x = y \\
  d(x,y) \leq q &\Longrightarrow d(y,x) \leq q \\
  d(x,y) \leq q \wedge d(y,z) \leq r &\Longrightarrow d(x,z) \leq q+r
\end{align*}
A \emph{Cauchy process} (or \emph{multi-valued modulated Cauchy sequence}) in a metric
space~$Y$ is a map~$\alpha : \QQ_{>0} \to P(Y)$ such that all the sets~$\alpha(q)$
are inhabited and such that~$d(x,x') \leq q + q'$ for all~$q,q' \in \QQ_{>0}$
and~$x \in \alpha(q), x' \in \alpha(q')$. Such a process \emph{converges
to}~$x_0 \in Y$ iff~$d(x,x_0) \leq q$ for all~$q \in \QQ_{>0}$
and~$x \in \alpha(q)$. A metric space~$Y$ is \emph{complete} iff every Cauchy
process in~$Y$ converges.

Verify, without using the law of excluded middle, that any complete metric
space is sober. \smallskip

\noindent{\scriptsize\emph{Note.} In the presense of the countable axiom of choice, any
Cauchy process can be refined to a Cauchy sequence. In its absence, however, the
familiar completion construction using equivalence classes of Cauchy sequences
fails to yield a Cauchy-complete space. In contrast, any Cauchy sequence and
more generally any Cauchy process converges in the space of equivalence classes
of Cauchy processes.\par}
\end{exercise}

\begin{exercise}{Isomorphisms of frames}%
\label{ex:isomorphisms-of-frames}%
Show that a frame homomorphism~$\alpha : A \to A'$ is an isomorphism (that is,
admits an inverse frame homomorphism) if and only if~$\alpha$ is surjective and
reflects the ordering (that is,~$\alpha(U) \preceq \alpha(V) \Rightarrow U
\preceq V$).
\end{exercise}

\begin{exercise}{The covariant approach to locales}%
Let~$A$ be a frame.
\begin{alphlist}[(c)]
\item Show that~$A$ contains arbitrary (set-indexed) meets (though they are not
required to be preserved by frame homomorphisms), by the impredicative
construction
\[ \bigwedge_{i \in I} U_i = \bigvee\{ V \in A \,|\, \text{$V \preceq U_i$ for all~$i \in I$}
\}. \]
\item What are the arbitrary meets in the frame of open subsets of a
topological space?
\item Show that for any frame homomorphism~$\alpha : A \to A'$ there exists a
monotone map~$\beta : A' \to A$ such that~\emph{$\beta$ is right adjoint
to~$\alpha$}, that is such that
\[ \alpha(U) \preceq V \quad\text{iff}\quad U \preceq \beta(V) \]
for all~$U \in A$ and~$V \in A'$.
\end{alphlist}
{\scriptsize\emph{Note.} Based on the observation in part~(c), a
\emph{covariant approach} to locales can be developed, in which locale maps~$X
\to X'$ are defined to be certain kinds of maps~$\O(X) \to
\O(X')$~\cite{picado-pultr:covariant}.\par}
\end{exercise}

@@

\begin{exercise}{Dense opens}%
\label{ex:dense-opens}%
An open~$U$ of a locale is \emph{dense} iff for any open~$V$, $U \wedge V =
\bot$ implies~$U = \bot$. 
\begin{alphlist}[(b)]
\item Let~$M$ be a set and endow it with the discrete topology. Show that an
open~$U \in \O(L(M))$ is dense iff for any element~$x \in M$, $\neg\neg(x \in
U)$.
\item Let~$\varphi$ be a truth value, hence an open of the one-point
locale~$\pt$. Show that~$\varphi$ is dense in~$\pt$ iff~$\neg\neg\varphi$.
\end{alphlist}
\end{exercise}

@@

\begin{exercise}{Coproduct of locales}%
\begin{alphlist}[(c)]
\item Show that the category of frames has small products.
\item Deduce that the category of locales has small coproducts.\smallskip

{\scriptsize\emph{Note.} The category of locales also has small products;
Exercise~\ref{ex:products-of-locales} is devoted to this fact.\par}
\item Let~$M$ be a set and endow it with the discrete topology. Show
that~$L(M)$ is the coproduct~$\coprod_{x \in M} \pt$.
\end{alphlist}
\end{exercise}

@@

\section{Geometric theories and sheaf models}

@@

\begin{definition}A \emph{geometric theory} consists of
\begin{enumerate}
  \item a set of sorts: $X$, $Y$, $Z$, \ldots
  \item a set of function symbols: $f : X \times Y \to Z$, \ldots
  \item a set of relation symbols: $R \hookrightarrow X \times Y \times Z$, \ldots
  \item a set of geometric sequents as axioms: $\varphi \vdash_{x:X, y:Y} \psi$, \ldots
\end{enumerate}
A \emph{geometric sequent} (in some context~$x_1\?X_1,\ldots,x_n\?X_n$) is a
formula built using only the ingredients ${=}\ {\top}\ {\wedge}\ {\bot}\
{\vee}\ {\bigvee}\ {\exists}$ and the relation symbols (but no ${\Rightarrow}\
{\forall}$). We follow the usual convention that negation is understood as
syntactic sugar for implication to~$\bot$. The symbol~``$\bigvee$'' refers to disjunction of arbitrary
set-indexed families of formulas. The theory \emph{proves} a sequent~$\sigma$ if and
only if there exists a \emph{derivation} of~$\sigma$, where the class of
derivations is inductively generated by the axioms and the rules displayed in
Table~\ref{table:geometric-logic}.\footnote{Because of the infinitary
disjunctions, some care is needed to not ``built in'' the axiom of choice in
the definition of derivations. More specifically, we allow inhabited sets of
subdeductions instead of requiring that subdeductions are given by a function.
Details on the implementation in~\textsc{czf} can be found in
Ref.~\cite[Definition~5.2]{rathjen:barr}.}
\end{definition}

@@

    (structural rules and rules for~${=}\ {\top}\ {\wedge}\ {\bot}\ {\vee}\ {\bigvee}\
    {\exists}$ as in Table~\ref{table:geometric-logic})
    \medskip

@@

Even though superficially similar, geometric theories play a
substantially different role than formal systems such as Peano arithmetic or
Zermelo--Fraenkel set theory. There are a number of notable differences:
\begin{enumerate}
\item Formal systems are typically of foundational interest and can be
fruitfully employed as metatheories. Geometric theories are more interesting
for applications, for instance for presenting spaces. This aspect is discussed
in Section~\ref{sect:presenting-frames}.
\item Formal systems typically come with an intended model; geometric theories
do not. For instance, the geometric theory of groups has all groups as models;
when we are writing down the axioms of a group, we are not setting out to
capture the properties of any one specific group.
\item Formal systems are typically required to be recursively
axiomatizable. Many interesting geometric theories are not, and indeed contain
an uncountable number of sorts, function symbols, relation symbols or axioms. Geometric theories also allow for arbitrary
set-indexed disjunctions; for this to make sense, the study of geometric
theories can only be carried out in a sufficiently rich background theory.
\item For formal systems, typically a version of Gödel's completeness theorem
holds: A formula is derivable iff it holds in all models. In contrast,
geometric theories can be consistent yet do not admit any set-based models. An
example is the geometric theory of surjections~$\NN \twoheadrightarrow \RR$
reviewed below.
\item Perhaps the most important difference to formal systems is that geometric
theories often depend on given mathematical objects. For instance, for each
ring~$A$, there is the geometric theory of prime ideals of~$A$. This theory
will play a crucial role in Section~\ref{sect:appl}.
\end{enumerate}

@@

In a similar vein, there is the geometric theory of monoids, of groups, and so
on. Some of these theories use the infinite disjunctions supported by geometric
logic: For instance, the theory of fields of positive characteristic or the
theory of torsion groups have among their axioms
\[ \top \vdash \bigvee_{n \in \NN_{\geq 1}} (\underbrace{1 + \cdots + 1}_{\text{$n$ summands}} = 0)
  \quad\text{or}\quad
  \top \seq{g:G} \bigvee_{n \in \NN} (g^n = e). \]

@@

\item \emph{The geometric theory of automorphisms of a field.}\label{par:theory-automorphisms}
Let~$L|k$ be a field
extension. The theory of automorphisms~$L \to L$ which fix~$k$ has
one nullary relation symbol~``$\sigma_{xy}$'' for each pair~$\langle x,y
\rangle \in L$ and the following axioms:
\begin{align*}
  \top &\vdash \bigvee_y \sigma_{xy} & \text{(for each~$x \in L$)} \\
  \top &\vdash \bigvee_x \sigma_{xy} & \text{(for each~$y \in L$)} \\
  \sigma_{xy} \wedge \sigma_{xy'} &\vdash \bigvee\{ \top \,|\, y = y' \} & \text{(for each~$x,y,y' \in L$)} \\
  \sigma_{xy} \wedge \sigma_{x'y} &\vdash \bigvee\{ \top \,|\, x = x' \} & \text{(for each~$x,x',y \in L$)} \\
  \top &\vdash \sigma_{00} \\
  \top &\vdash \sigma_{11} \\
  \sigma_{xy} \wedge \sigma_{x'y'} &\vdash \sigma_{x+x',y+y'} & \text{(for each~$x,x',y,y' \in L$)} \\
  \sigma_{xy} \wedge \sigma_{x'y'} &\vdash \sigma_{xx',yy'} & \text{(for each~$x,x',y,y' \in L$)} \\
  \top &\vdash \sigma_{xx} & \text{(for each~$x \in k$)}
\end{align*}
A model of this theory (can be reorganized to be) an element of the classical
Galois group of~$L|k$.

@@

\subsection{Sheaves on locales}
\label{sect:sheaves}

@@

In some cases, these new techniques XXX
